Imperative Programming with Variational Effects

### 2018-08-06 paper/paper.tex
work on introduction

%% TODO
Variation is a constant element in modern software systems.

>>>>

Variation is an extremely common element in modern software. Variation in software can be
thought of as broadly belonging to two interrelated categories: software systems and variational execution.
Variation in software systems is commonly found in preprocessor directives, command line options, or aspect- or feature-oriented
programming languages (TODO: cite aspect and feature oriented work here).
These all share the common behavior of taking some set of features as input and
producing a program or executable uniquely determined by those features as output. Variational execution occurs whenever a single
program is run multiple times on related inputs. Such serial execution of the same program over different inputs can be viewed in the abstract
as equivalent to a single execution across a value representing all inputs and producing a similar value representing all outputs.
The key distinction between these two views of variation is that one takes a set of features and produces programs or executables that
operate over normal non-variational values, while the other takes a variational set of values and a single program and produces a corresponding variational set of outputs.

While seemingly distinct, there is considerable overlap between these two areas of variation. Tooling or static analysis that deals with
variation software systems often encounters the same problems and employs the same techniques to solve them
that are found in variational execution. Chief among these issues is that brute-force techniques for dealing with variation quickly become infeasible as the
number of features grows. For a variational software system with $n$ features, the number of unique variants produced is $2^n$. Similarly, the naive approach
of running a program multiple times over varying inputs requires $2^n$ executions for inputs with $n$ features. This exponential growth poses what appears to
be an insurmountable obstacle to efficiently handling variation in the face of a large number of distinct features.

Of course, the situation is not quite as hopeless as it might first appear. In both variational software systems and variational models of execution
there are significant elements that are shared in common between variants. In software systems this might take the form of pieces of code such as functions
or classes that are shared between multiple variants. In the case of variational execution this means computations that are performed across several variants.
By exploiting these commonalities what previously seemed impossible often becomes quite efficient. Software systems analysis that takes advantage of variation
in this way---commonly termed \emph{variationally-aware} analysis---has been developed across a number of areas, including typechecking, type inference, and
testing (TODO: citations). In turn, and often in support of such variationally-aware analysis tools, systems of variational execution have been developed for a number
of languages, including PHP, Java, and Javascript (TODO: citations), including the development of efficient variational data structures to support such execution (TODO: citations).
